---
title: "Seminar 7"
author: "Natalia Lamberova"
date: "11/10/2016"
output:
  pdf_document: default
  html_document: default
---


## Final Project Info

***Full draft*** including DeclareDesign is due on ***November 23***     
***Presentations: November 30 and December 2***    
***Final Draft of Registered Report: December 6***    
***Final Draft of the analysis: December 16***


## Important Dates

Task                      | Date        | Time
--------------------------|-------------|------
Research Question         | November 18 |
Design in Words           | November 18 | 6 pm
Measurement Strategy      | November 23 | 12 pm
DeclareDesign & Diagnosis | November 23 | 12 pm
Ethical Issues            | November 23 | 12 pm
Registered report         | December 6  | 6 pm
***Fake data delivered*** | December 7  |
Results                   | December 16 | 6 pm
Discussion                | December 16 | 6 pm

## Setting up the design in Declare Design

```{r}
library(pacman)
p_load(DeclareDesign,knitr)
#Here we've loaded the needed packages and setting the seed
set.seed(1234)
```

## Data Simulations:


## Data Simulations: Multivariate:
```{r}
multivariate_pop <- declare_population(
  income = declare_variable(type = "normal",
                            location_scale = c(mean = 1000, sd = 10)),
  gender = declare_variable(type = "gender"),
  party = declare_variable(type = "us_party"),
  employed = declare_variable(type = "binary"),
  size = 500
)

multivariate_draw <- draw_population(population = multivariate_pop)
head(multivariate_draw)
```
## Data Simulations: Linear Otcomes
If you wish to simply create linear potential outcomes, you must specify the different conditions, and include a function that specifies how they are generated in a linear model.
```{r}
# Make a very simple sample frame
simple_frame <- declare_population(one_level = list(),size = 10)
# Make a very simple assignment scheme
condition_names <- c(0,1)
assignment <- declare_assignment(condition_names = condition_names)
# Make the draw
pop_draw <- draw_population(population = simple_frame)
pop_draw <- assign_treatment(data = pop_draw,
                                assignment = assignment)

outcomes_1 <- declare_potential_outcomes(
  # Specify the conditions of the experiment
  condition_names = condition_names,
  # Here the average treatment effect is .5
  formula = outcome ~ .5*Z 
    )


linear_draw <- draw_outcome(data = pop_draw,
                            potential_outcomes = outcomes_1)
head(linear_draw)
```
## Non-linear Outcomes:
```{r}
# For a probit data-generating process, write a predictive probit function  
probit_outcome <- function(Z){
  rbinom(n = length(Z), size = 1, prob = pnorm(q = Z*.2))
}
condition_names <- c(0,1)
# Insert the function into the outcome formula
probit_PO <- declare_potential_outcomes(
  condition_names = condition_names,
  formula = Y ~ probit_outcome(Z)
)

# Make data
probit_draw <- draw_outcome(data = pop_draw,
                            potential_outcomes = probit_PO)
head(probit_draw)
```
## Poisson Process:
Think about witing for the bus. $\lambda$ is the event rate (average number of events per interval).
So the Poisson process look like this:
$$P(kEventsInInterval)=\frac {\lambda e^{-\lambda}} {k!}$$





```{r}
# For a Poisson data-generating process, write a predictive poisson function  
poisson_outcome <- function(Z){
  rpois(n = length(Z), lambda = exp(Z*.2))
}

# Insert the function into the outcome formula
pois_PO <- declare_potential_outcomes(
  condition_names = condition_names,
  formula = Y ~ poisson_outcome(Z)
)

pois_draw <- draw_outcome(data = pop_draw,
                            potential_outcomes = pois_PO)
head(pois_draw)
```
##Changes in proportions:
Simple Changes in Proportions
You may want to specify potential outcomes at the group-level, especially when the outcome is binary or categorical. For example, we might want to specify that the treatment increases uptake of a service by 5%, or we might want to specify that it increases the probability of voting for candidate A by 10%, while reducing that of voting for B and C by 6% and 4%, respectively. Binary case in presented below:

```{r}
success_proportions <- c(control = .3, treatment = .8)

proportion_PO_1 <- declare_potential_outcomes(
  potential_outcomes_function = proportion_potential_outcomes_function,
  condition_names = condition_names,
  assignment_variable_name = "Z",
  outcome_variable_name = "Y",
  population_proportions = success_proportions)

prop_draw_1 <- draw_outcome(data = pop_draw,
                            potential_outcomes = proportion_PO_1)
head(prop_draw_1)
```
## Multilevel:
```{r}
N_per_level <- c(students = 8,
                 classrooms = 4,
                 schools = 2)

# Note: level names are inhereted from the variable declarations, not 
# from the the N_per_level vector

school_pop <- declare_population(
 student = list(
   math_score = declare_variable(type = "normal",
     location_scale = c(10,5)),
   lang_score = declare_variable(type = "normal",
     location_scale = c(10,5))
 ),
 classroom = list(
   N_students_raw = declare_variable(
     type = "normal",
     location_scale = c(20,1)),
   N_students = "round(N_students_raw,0)"
 ),
 school = list(
   poor_area = declare_variable(type = "binary")
 ),
 size = N_per_level
)

school_draw <- draw_population(population = school_pop)
head(school_draw)
```
## A Six-Step Workflow for Experiments with Custom Functions
## Custom population function just selects N random numbers.
A population function has to create a data frame whenever it is run. It can optionally have inputs for N, N_per_level, or group_sizes_per_level (but not more than one) to indicate the sample size. In this example, the input N is used and set in the declaration to 100 (note this allows you to reuse my_pop_function to test out designs with differing sample sizes).
```{r}
N<-100
my_population <- function(size) { data.frame(income = rnorm(size)) }
population      <- declare_population(
  custom_population_function = my_population, 
  size = N)
```
Declaring potential outcomes requires that you declare both the ***treatment levels*** and a ***potential outcomes*** formula,ie a statement that maps from possible treatment conditions to Ys.It must indicate what value Y takes for all values that a treatment might take.

This formula and the condition names defined in the declaration place restrictions on the remaining workflow, by setting the outcome name to Y and the names of the conditions. When potential outcomes are constructed in mock data, they will be named in this case Y_Z1 and Y_Z0. When realized outcomes based on a treatment assignment are created, they will be named in this case Y.
```

## Potential Outcomes:
```{r}
my_potential_outcomes <- function(data) { 
  (data$Z == "Z1") * 0.25 + runif(nrow(data)) }
potential_outcomes <- declare_potential_outcomes(
  condition_names = c("Z0", "Z1"), 
  potential_outcomes_function = my_potential_outcomes,
  outcome_variable_name = "Y")
```

## A sampling function then takes a data frame and indicates which rows to select.
```{r}
# Taking half the units.

my_sampling    <- function(data) {
  N <- nrow(data); n <- floor(N/2); 
  sample(rep(c(0, 1), each = N-n), N) }
sampling       <- declare_sampling(
  custom_sampling_function = my_sampling)
```

## An estimand function takes a data frame with potential outcome columns and returns a scalar.
```{r}
#Estimand - Difference in means

my_estimand    <- function(data) { 
  mean(data$Y_Z_Z1 - data$Y_Z_Z0) }
estimand       <- declare_estimand(
  estimand_function = my_estimand, 
  potential_outcomes = potential_outcomes)
```

## An assignment function takes a data frame and returns a vector of treatment assignments
```{r}
my_assignment  <- function(data) { 
  N <- nrow(data); 
  sample(c("Z0", "Z1"), 
         N, replace = T) }
assignment     <- declare_assignment(
  custom_assignment_function = my_assignment, 
  potential_outcomes = potential_outcomes)
```

##An estimates function 
takes data and returns a matrix with columns representing estimates and rows representing statistics of the estimates, such as the estimate itself, the standard error, the p-value, etc.

```{r}


my_estimates   <- function(data) { 
  est      <- mean(data$Y[data$Z == "Z1"]) 
  - mean(data$Y[data$Z == "Z0"])
  se       <- sqrt(var(
  data$Y[data$Z == "Z1"])/sum(data$Z == "Z1") + 
  var(data$Y[data$Z == "Z0"])/sum(data$Z == "Z0"))
  df       <- nrow(data) - 2
  p        <- 2 * pt(abs(est/se), df = df, lower.tail = FALSE)
  ci_lower <- est - 1.96*se
  ci_upper <- est + 1.96*se
  data.frame(estimate_label = "diff-in-means",
             est = est, se = se, p = p, 
             ci_lower = ci_lower, ci_upper = ci_upper, df = df,
             stringsAsFactors = FALSE)
}
estimator      <- declare_estimator(
  estimates = my_estimates, 
  estimand = estimand)
```

## Full design in six steps.
```{r}
my_design      <- declare_design(
  population = population, 
  sampling = sampling, 
  potential_outcomes = potential_outcomes,
  assignment = assignment, 
  estimator = estimator, 
  label = "my-design")
```


## What next?
```{r}
#Constructing mock data
mock_population <- draw_population(
  population = population, 
  potential_outcomes = potential_outcomes)
mock_sample <- draw_sample(
  data = mock_population, 
  sampling = sampling)
mock_sample <- assign_treatment(
  data = mock_sample, 
  assignment = assignment)
mock_sample <- draw_outcome(
  data = mock_sample, 
  potential_outcomes = potential_outcomes)
kable(head(mock_sample), digits = 3, 
      row.names = FALSE)
```

## Diagnosis

You can define a custom statistic function which takes as options estimates, estimators, or both and returns a scalar summary of them. Here, we calculate the proportion of estimates that are larger than zero.
```{r}

my_statistic <- function(estimates, ...) { 
  mean(sapply(1:length(estimates), function(i)
  as.numeric(estimates[[i]]["p", , drop = FALSE])) > 0.5) }

diagnosis <- diagnose_design(design = my_design, 
                             population_draws = 10, 
                             sample_draws = 10)

kable(summary(diagnosis, statistics = list(
  calculate_PATE, calculate_sd_SATE, calculate_power, 
  calculate_RMSE, calculate_bias, calculate_coverage, 
  my_statistic), labels = c("PATE", "sd(SATE)", "power", 
                            "RMSE", "bias", "coverage",
                           "custom_stat")), digits = 3)
```


## Interpreting Diagnosands

***Bias***        
*If we ran the experiment again and again, will the average of our estimates exactly equal our estimand?*    

$${\rm bias} \equiv E(\hat\tau - \tau)$$
***Type S error***   
"the probability that the estimate has the incorrect sign, if it is statistically significantly different from zero"

Note this interprets an estimate of $\tau$ that is not statistically significant as not an exaggeration

\href{http://www.stat.columbia.edu/~gelman/research/unpublished/retropower.pdf}{Gelman and Carlin, "Beyond Power Calculations"}

***RMSE***       
*How noisy is my estimate of the estimand?*

$${\rm RMSE} \equiv \sqrt{E( {\hat\tau - \tau)}^2 }$$

***Coverage***    
*Are the standard errors a good estimate of the uncertainty of my estimates, or are they too conservative or liberal?*

$$E({\rm CI}_- \leq \tau \leq {\rm CI}_+) \rightarrow^p 1 - \alpha$$
"Nominal coverage" for $\alpha = .05$ is $0.95$: the confidence interval should contain the estimand 95 out of 100 times the experiment is run

When the coverage probability is below $.95$, the confidence intervals are too narrow (i.e. the standard errors are too small).

***Power***     
*How often will I reject the null hypothesis when it is false?*

Reference value of power $= 0.8$


## Simulating results
```{r}
mock_population <- draw_population(population = population)
mock_population <- assign_treatment(
  data = mock_population, 
  assignment = assignment)
mock_population <- draw_outcome(data = 
                                  mock_population, 
                                potential_outcomes = 
                                  potential_outcomes)
kable(get_estimands(estimator = estimator, 
              data = mock_population), digits = 3)
```

## Communicating results
```{r}
kable(get_estimates(estimator = estimator, data = mock_population), digits = 3)
```


